Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-08-28 SPH-Torus-paper1/sphtorus_benchmark.tex
Minor updates.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2027 25464f21-6099-402c-b30a-640f1d40d38b

There is not a unique approach to coupling an SPH code to a radiative
transfer code and a number of different approaches have been developed
(\cite{nayakshin_2009}, \cite{pawlik_2008} and references
therein). This paper focuses on radiative transfer calculations using
an adaptive grid in the context of modelling processes in star
formation.

Accurately coupling the SPH component to the radiative transfer
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
cicumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

Two grid splitting conditions were tested. The first method splits a
grid cell if the mass within the cell exceeds a given limit. For this
condition the resolution is highest where the density is highest and
the grid cell size is analogous to the SPH particle smoothing
length. The second condition decides whether to split the cell based
on the fractional density difference between the most dense and least
dense SPH particles contained within the cell. The quantity
\begin{equation}
f_{\rm{split}}=\frac{\rho_{\rm{max}} - \rho_{\rm{min}}}{\rho_{\rm{max}} + \rho_{\rm{min}}}
\end{equation}
is calculated, where $\rho_{\rm{max}}$ is the highest SPH particle
density and $\rho_{\rm{min}}$ is the lowest SPH particle density, and
the cell is split if a specified value of $f_{\rm{split}}$ is
exceeded. The first condition gives a resolution like that of the
original SPH representation whereas the second condition should be
better at representing gradients in density which are likely to be
important for radiative transfer calculations.

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc.
A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years). The freqency
with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc.

@@

The surface density of the disc is shown in Fig.~\ref{fig:sigma} where
the solid line is the vertically isothermal disc shown in
Fig.~\ref{fig:frame1} and the dotted line is the disc at the final
time step (Fig.~\ref{fig:frame6}).  Away from the inner and outer
edges of the disc the surface density is well represented by a power
law profile as a function of radius.
\begin{figure}
  \includegraphics[scale=0.3]{figures/sigma}
  \caption{Surface density as a function of radius.}
  \label{fig:sigma}
\end{figure}
The scale height of the disc was determined by fitting Gaussians to
the vertical density profile in 100 radial bins. The fitted scale
height, as a function of radius, is plotted in
Fig.~\ref{fig:scale_height}.  Within the central 10~AU the density
profile appears flat so it was not possible to fit a meaningful scale
height. The solid line is the isothermal disc and the dotted line is
the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/scale_height}
  \caption{Disc scale height as a function of radius.}
  \label{fig:scale_height}
\end{figure}
The average temperature within one scale height of the disc mid plane
is plotted in Fig.~\ref{fig:mid_tem}. The solid line is the isothermal
disc and the dotted line is the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/mid_tem}
  \caption{Average temperature within one scale height of the mid plane as a function of
    radius.}
  \label{fig:mid_tem}
\end{figure}

@@

We have presented results from a combined SPH and grid-based radiative
transfer code, which has been used to model a non-self gravitating
circumstellar disc. Validation tests of the method have revealed a
number of factors which must be taken into consideration when setting
up a model of this type.

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SED. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a managable
level with a realistic number of SPH particles in the disc.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. As the disc settles towards an equilibrium state, vertical
motions are seen which give the appearance of an outward propagating
scale height fluctuation. This is a vertical motion which decays more
rapidly at smaller radii and is seen whether radiative transfer is
included or not. One consequence of these fluctuations is that they
can intecept radiation from the central star which can then be
vertically transported into the mid-plane of the disc. Even thought
the fluctuations are of a small amplitude they exert an influence on
the mid-plane temperature.

>>>>

There is not a unique approach to coupling an SPH code to a radiative
transfer code and a number of different approaches have been developed
(\cite{pawlik_2008} and references therein). This paper focuses on
radiative transfer calculations using an adaptive grid in the context
of modelling a circumstellar disc.

Accurately coupling the radiative transfer component to the SPH
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
cicumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

Two grid splitting conditions were tested. The first method splits a
grid cell if the mass within the cell exceeds a given limit. For this
condition the resolution is highest where the density is highest and
the grid cell size is analogous to the SPH particle smoothing
length. The second condition decides whether to split the cell based
on the fractional density difference between the most dense and least
dense SPH particles contained within the cell. The quantity
\begin{equation}
f_{\rm{split}}=\frac{\rho_{\rm{max}} - \rho_{\rm{min}}}{\rho_{\rm{max}} + \rho_{\rm{min}}}
\end{equation}
is calculated, where $\rho_{\rm{max}}$ is the highest SPH particle
density and $\rho_{\rm{min}}$ is the lowest SPH particle density, and
the cell is split if a specified value of $f_{\rm{split}}$ is
exceeded. The first condition gives a resolution like that of the
original SPH representation whereas the second condition should better
represent gradients in density which are important for radiative
transfer calculations.

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc
(i.e. the ``prompt escape'' regime decribed by \cite{nayakshin_2009}).
A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years). The freqency
with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc.

@@

The surface density of the disc is shown in Fig.~\ref{fig:sigma}, the
scale height is shown in Fig.~\ref{fig:scale_height} and the radial
temperature profile is shown in Fig.~\ref{fig:mid_tem}.  The solid
lines are from the vertically isothermal disc shown in
Fig.~\ref{fig:frame1} and the dotted lines are the disc at the final
time step shown in Fig.~\ref{fig:frame6}. The scale height of the disc
was determined by allocating particles to 100 radial bins and fitting
a Gaussian to the vertical density profile. The radial temperature
profile is the average temperature within one scale height of the
mid-plane for each radial bin.
\begin{figure}
  \includegraphics[scale=0.3]{figures/sigma}
  \caption{Surface density as a function of radius.}
  \label{fig:sigma}
\end{figure}
\begin{figure}
  \includegraphics[scale=0.3]{figures/scale_height}
  \caption{Disc scale height as a function of radius.}
  \label{fig:scale_height}
\end{figure}
\begin{figure}
  \includegraphics[scale=0.3]{figures/mid_tem}
  \caption{Average temperature within one scale height of the mid plane as a function of
    radius.}
  \label{fig:mid_tem}
\end{figure}     
The majority of the surface density profile is well represented by a
power law and does not change significantly as a result of including
radiative transfer, except within the central
region. Figures~\ref{fig:scale_height} and \ref{fig:mid_tem} show that
the disc is hotter close to the mid-plane, except within a region
between 10--20 AU of the centre, but has a smaller scale
height. Figure~\ref{fig:log_u} shows that the vertical temperature
gradient in the disc is such that temperature increases with height
out of the mid-plane. Although the disc is hotter when radiative
transfer is included the vertical temperature gradient tends to reduce
the pressure gradient which is required for hydrostatic support
resulting in a smaller scale height.

@@

We have presented results from a combined SPH and AMR grid-based
radiative transfer code, which has been used to model a non-self
gravitating circumstellar disc. Validation tests of the method have
revealed a number of factors which must be taken into consideration
when setting up a model of this type.

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SEDs. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a managable
level with a realistic number of SPH particles in the disc.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. As the disc settles towards an equilibrium state,
vertical motions are seen which give the appearance of an outward
propagating scale height fluctuation. This is a vertical motion which
decays more rapidly at smaller radii and is seen whether radiative
transfer is included or not. These are not the same as the
instabilities seen in other radiation hydrodynamics disc models
\citep{min_2009}. One consequence of these fluctuations is that they
can intecept radiation from the central star which can then be
vertically transported into the mid-plane of the disc hence even
thought the fluctuations are of a small amplitude they exert an
influence on the mid-plane temperature.
